---
title: "pTITAN2"
subtitle: "occurrences and permute"
output:
  bookdown::html_document2: default
bibliography: references.bib
vignette: >
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteIndexEntry{pTITAN2-occurrences}
 %\VignetteEncoding{UTF-8}
---

```{r label=setup, include = FALSE}
library(knitr)
knitr::opts_chunk$set(collapse = TRUE)
```

**Objective:** Provide extended details on the `occurrences` function.  This
includes the how and why of the function and examples reproducing the output of
the `occurrences` function using different R data manipulation dialects: base R,
the [Tidyverse](https://www.tidyverse.org/), and
[data.table](https://rdatatable.gitlab.io/data.table/).

# Introduction

The `occurrences` function selects the codes that should be used for the
`TITAN2::titan` run. The goal is to select the macroinvertebrate code with the
most taxonomic detail having at least `n` occurrences. Only one
macroinvertebrate code will be associated with the macroinvertebrate counts. For
example, if there are at least `r as.list(args(pTITAN2::occurrences))$n`
occurrences at the genus level, the family, order, and class codes would not be
used in the `TITAN2::titan` run.

The names within the input taxonomic data set are expected to have the following
structure:

* 8 characters in length
* characters 1 and 2 denote the class
* characters 3 and 4 denote the order
* characters 5 and 6 denote the family
* characters 7 and 8 denote the genus.

If no information at a level exists, use "00" to hold the place. For example: A
code that is 'Bi000000' is the Bivalvia class, while BiVe0000 is the Bivalvia
class, Veneroida order. BiVeSh00 is the Bivalvia class, Veneroida order,
Spheriridae family. BiVeSh01 is a genus within that family.

Taking taxonomic data as an input, the return of `occurrences` function is a
`data.frame` with the taxon, the class, order, family, and genus split out into
individual columns, and the count of occurrences within the provided taxonomic
data set.

`TITAN2::titan` recommends all taxonomic groups have at least five observations
[@baker2010new]. Thus, `occurances` returns only taxons with at least `n`
observations, defaulting to
`r as.list(args(pTITAN2::occurrences))$n`.
The taxonomic code chosen for analysis should be at the finest possible
resolution.  For example, if a macroinvertebrate count has at least
`r as.list(args(pTITAN2::occurrences))$n` occurrences in a
genus code, the family, order, and class codes associated with these counts
should be removed. Further, if there are too few counts at the genus level, but
at least 
`r as.list(args(pTITAN2::occurrences))$n` counts at the family level- the
family code would be retained and the order and class codes would be removed.



# Example

To reproduce the example in this vignette you will need to load and attach
the pTITAN2 and magrittr namespaces.  Other namespaces are used explicitly,
loaded (not attached) here.

```{r }
library(pTITAN2)
library(magrittr)
loadNamespace("data.table")
loadNamespace("dplyr")
loadNamespace("tidyr")
```

The taxonomic sets, CD_06_Mall_wID and CN_06_Mall_wID, contains raw
macroinvertebrate density data for 500 possible macroinvertebrate codes for
each taxonomic level (class, order, family, genus).
```{r}
dim(CD_06_Mall_wID)  # Taxonomic, Dry Treatment
dim(CN_06_Mall_wID)  # Taxonomic, Normal Treatment

# top 4 rows and first 10 columns
CD_06_Mall_wID[1:4, 1:10]
CN_06_Mall_wID[1:4, 1:10]
```

```{r}
occurrences
```

The function `occurrences` is an extension for
deciding the taxonomic detail to be included in a TITAN run based on `minSplt`
option in `TITAN2::titan()`.
`minSplt` is minimum number of occurrences that TITAN is looking for
taxa to have across the provided sites. The `minSplt` default in TITAN is 5 and
should never drop below 3.  The default for `occurrences` is `minSplt` = `n` =
`r as.list(args(pTITAN2::occurrences))$n`.

```{r label = "occurrences_example"}
head(occurrences(CN_06_Mall_wID[, -1]))
```

Compare these results to working with the raw data.  For example purposes we
present the summary of the raw data twice,
once using tidyverse syntax and once using data.table syntax.
```{r}
# Tidyverse
CN_06_Mall_wID %>%
  dplyr::select(-StationID) %>%
  tidyr::pivot_longer(
                      cols = dplyr::everything(),
                      names_to = 'taxon',
                      values_to = 'count'
                      ) %>%
  dplyr::mutate(
                Class  = substr(.data$taxon, 1L, 2L),
                Order  = substr(.data$taxon, 3L, 4L),
                Family = substr(.data$taxon, 5L, 6L),
                Genus  = substr(.data$taxon, 7L, 8L)
                ) %>%
  dplyr::group_by(.data$Class, .data$Order, .data$Family, .data$Genus) %>%
  dplyr::summarise(
                   taxon = unique(.data$taxon),
                   count = sum(.data$count > 0),
                   .groups = "keep"
                   ) %>%
  dplyr::ungroup() %>%
  dplyr::filter(.data$count >= 5) %>%
  dplyr::arrange(.data$Class, .data$Order, .data$Family, .data$Genus)
```

Using data.table:
```{r}
# data.table
taxon_count <- data.table::copy(CN_06_Mall_wID)
data.table::setDT(taxon_count)
data.table::set(taxon_count, j = "StationID", value = NULL)
for(j in as.integer(which(sapply(taxon_count, is.integer)))) {
  data.table::set(taxon_count, j = j, value = as.numeric(taxon_count[[j]]))
}

taxon_count <- data.table::melt(taxon_count,
                                variable.factor = FALSE,
                                measure.vars = names(taxon_count),
                                variable.name = "taxon",
                                value.name = "count")
taxon_count[, count := sum(count > 0), keyby = .(taxon)]
taxon_count <- unique(taxon_count)

data.table::set(taxon_count, j = "Class",  value = substr(taxon_count[["taxon"]], 1L, 2L))
data.table::set(taxon_count, j = "Order",  value = substr(taxon_count[["taxon"]], 3L, 4L))
data.table::set(taxon_count, j = "Family", value = substr(taxon_count[["taxon"]], 5L, 6L))
data.table::set(taxon_count, j = "Genus",  value = substr(taxon_count[["taxon"]], 7L, 8L))

taxon_count <- subset(taxon_count, taxon_count$count >= 5)

data.table::setkeyv(taxon_count, c("taxon", "Class", "Order", "Family", "Genus"))

taxon_count
```

Note that for the Ar class there is only one row with no order, family, or genus
level information.  Compare to the Bi class where the Un order has no presence
counts and is thus not reported in object returned from `occurrences`.  BiVeCa01
has counts and will be reported but BiVeCa00 should not be reported.  BiVe0000
and Bi000000 should not be reported as `occurrences` as preference for the codes
with family and genus level information.

## permute

The function permute is used to generate a list of permuted sets of taxa and
environmental gradients. Function parameters include a list of data frames
containing taxa for each treatment group, a list of data frames containing the
associated environmental gradient variables, and the site ids. Before we can run
permute, we need to import the environmental gradients data.

```{r }
eg <-
  permute(taxa = list(CD_06_Mall_wID, CN_06_Mall_wID),
          envs = list(C_IC_D_06_wID, C_IC_N_06_wID),
          sid  = "StationID")
str(eg, max.level = 2)
```

The return of `permute` is list of lists. The first level denotes the treatment,
in this example Treatment1 is "dry" and Treatment2 is "normal" -- the order of
the input data sets.  The second level contains the data.frames with
environmental and taxonomic data.


# References
<div id="refs"></div>

